Fitting metabolic models to dissolved oxygen data: The Estuarine BAyesian Single-station Estimation method (EBASE)

### 2024-02-27 revision/review_response.qmd

---
format: 
  docx:
    reference-doc: response_styles.docx
bibliography: ../refs.bib

execute: 
  echo: false
  warning: false
  message: false
---

```{r echo = F, cache = F, eval = F}
#| eval: false
spelling::spell_check_files(here::here('revision/review-response.qmd'))
```

*We thank all three reviewers for providing constructive comments that have helped to improve the manuscript. Our responses to each of these comments are provided below.  Note that line number refer to those in the original manuscript.*

@@

Line 32-36: Estuaries are also significantly influenced by their freshwater end-members, consider adding ‘river’ to ‘…interface between land, river, ocean, and atmosphere’. Similarly consider incorporating rivers into ‘…transported from land to rivers to ocean…’.

*Response: Added.*

Line 47: Consider adding how materials are lost to the atmosphere, the sediments, and the ocean. For example, “…that are lost to the atmosphere (i.e., outgassing), the sediments (i.e., burial), and the ocean (i.e., advection).”.

Line 48-50: Consider adding the underwater eddy covariance method to the list of methods for estimating aquatic metabolism. See Berg et al. 2022 (https://doi.org/10.1146/annurev-marine-042121-012329)

*Response: Added to the list, including the citation, "...use of inert gases, and aquatic eddy covariance (Kemp and Testa 2011; Staehr et al. 2011; Berg et al. 2022).*

@@

Line 80: As written, it seems ‘assumptions’ should be singular, not plural (…can violate an assumption of the method…).

*Response: Corrected.*

Line 82: I appreciate that the sign of P and R estimates in the literature has been applied non- uniformly. However, I believe the convention is that P is positive (DO is produced) while R is negative (DO is consumed). Could you change the physically impossible statement to state negative P and positive R? Or add a sentence about why both values are described with a positive sign – this is somewhat described on line 236.

*Response: Yes, we agree the sign has not been consistently described in the literature. We have added clarifying text at the end of the paragraph: "Note that R is often expressed as a negative value indicating consumption of dissolved oxygen.  Here and throughout, we express correct values of R as positive values given the metabolic model described below."*

@@

Line 106: Curious why your method builds off BASE when there has been an updated Bayesian technique for stream metabolism that more accurately estimates uncertainty and reduces equifinality - please see Appling et al. (2018) cited above. Additionally, one of the sources cited above (Tassone and Bukaveckas 2019) used the updated Bayesian technique to estimate single- station estuarine metabolism.

*Response: text goes here*

Line 126: Presumably, EBASE produces daily metabolic estimates. What constitutes a day in EBASE? Is it from midnight to midnight the following day or from 4 am to 4 am?

*Response: The "metabolic" day in EBASE is midnight to midnight the following day. Clarifying text was added to line 182.  In our experience, very similar estimates are obtained whether the day is defined midnight to midnight or some other period.  This is especially true if the optimization period is greater than one day, as suggested in model application, i.e., lines 317 - 320, 523-525.*

@@

Line 129-131: I normally see water column depth represented as Z not H. Should clarify if the values of P, R, and D are positive, negative or both.

*Response: The water column depth notation was changed to Z (including the documentation and code in EBASE). Clarification was added for P, R, and D on line 132: "Both $P$ and $R$ are expressed as positive values, whereas D can be positive (outgassing) or negative (ingassing)."*

@@

Line 135: Consider making C = Cobs if it represents the observed DO concentration. If that is not what it represents, then will need to be described in text.

*Response: Yes, C represents observed dissolved oxygen concentration (as noted on line 132).  We have retained C instead of using Cobs for simplicity in the model equations.  Further, the addition of the model input/output table as requested by another reviewer helps clarify the input data.*

@@

Line 181: Specifying the model optimization period is an advancement of the BASE model but is not described in text. What does ‘optimization’ mean? What are the implications of using 1 day, 7 days, 30 days, or the entire time series length on the output? If you use 7 days, do you only get weekly estimates of metabolism?

*Response: This was clarified in the text.*

*Line 181:  "First, users can specify the model optimization period as the amount of time over which the unknown parameters are estimated from the Bayesian prior distributions (i.e., $a$, $R$, and $b$, see below)."*

*Line 184: "Note that only one estimate for each of $a$, $R$, and $b$ is returned for each optimization period as defined by the model, although $P$ and $D$ can vary continuously with each timestep given the dependence on other inputs that can vary across the time series (e.g., $PAR$ and wind)."*

@@

Line 185: On line 546 it is stated that linear interpolation is used, that should be described here. Furthermore, the threshold discussion on line 547-551 should be mentioned here. Lastly, on line 550-551, a 12 hour threshold should be further discussed. Why 12- hours? Why not 6 considering anything longer than 6 and you would be missing half of a semidiurnal tidal cycle.

*Response: 12 hours was chosen as a default value to cover approximately half of a diurnal cycle.  Users may want to consider smaller gaps if tidal cycles are having a large influence on the observed time series.  Our default value was chosen under the assumption that tidal effects, although they may be present, are likely not the dominant signal in the data.  If the latter is expected, detiding is strongly encouraged.  We have added text to this effect.*

*Line 186: "All gaps are linearly interpolated to allow the model estimation to proceed continuously and any results for an optimization period with gaps larger than 12 hours are removed, eliminating the need to pre- and post-process the data when using EBASE. The user can vary the criteria for removing results (e.g., 6 hours as a more conservative criteria)."*

*Line 555: "More conservative gaps can be chosen if, for example, semidiurnal tidal signals operating at time scales less than 12 hours are expected to influence the results.  However, detiding is strongly encouraged in these cases and the default gap size of 12 hours that describes the diurnal oxygen signal is likely appropriate after detiding."*

*Also note that data for an entire optimization period are removed if any gaps greater than the chosen gap size are present, providing an additional safety measure that the results are not based on inaccurately interpolated inputs.*

### 2024-02-27 manu-draft.qmd
more revisions

```{=tex}
\newcommand{\auni}{\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)}
\newcommand{\runi}{\text{mmol}/\text{m}^2/\text{d}}
\newcommand{\buni}{\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)}
```

*Running head*: EBASE metabolic model

@@

Estuaries play a key role in the global cycling of elements because of high rates of biogeochemical activity at the interface between land, ocean, and atmosphere. As materials are transported from land to ocean, estuaries profoundly transform or filter these materials through various processes, chief among which are primary production and respiration, collectively referred to as metabolism [@schubel1984]. For example, estuarine respiration of organic carbon delivered from land produces carbon dioxide, which may be lost to the atmosphere via outgassing [e.g, @cai2011]. Similarly, estuarine primary production consumes dissolved inorganic carbon delivered from land, forming particulate organic carbon that may be buried in estuarine sediments [e.g., @hu2006]. Estuarine metabolic processes are fundamentally important, as illustrated by the dissolved inorganic carbon budget of eastern North American estuaries [@najjar2018]; net primary production and heterotrophic respiration far exceed inputs from upland sources and outputs to the atmosphere and ocean. As such, small changes in either net primary production or heterotrophic respiration can have a large impact on the difference between the two, net ecosystem metabolism (NEM, sometimes also referred to as net ecosystem production), which ultimately governs the form and quantity of estuarine materials that are lost to the atmosphere, the sediments, and the ocean.

Methods for estimating metabolism in aquatic environments are numerous and include open-water techniques, bottle-based incubations, ecosystem budgets, oxygen isotopes, and use of inert gases [@kemp2011; @staehr2011]. Based on the variety of assumptions and limitations of each method, comparisons of metabolic rates within and between ecosystems have been challenging. For example, an assessment in Randers Fjord (Denmark) showed agreement in both the sign and magnitude among four different approaches [@gazeau2005]. However, gross primary production and ecosystem respiration estimates from incubations were consistently lower than those from open-water methods, consistent with the estuarine respiration synthesis of @hopkinson2005. Even within one technique, there are numerous choices to be made that alter the outcome. For example, in the commonly used ${}^{14}$C technique for measuring primary production, @cloern2014 noted that investigators have made divergent choices with regard to filtering mesozooplankton grazers, measuring ${}^{14}$C in dissolved organic carbon, changing spectral light quality with depth, the euphotic zone depth, the incubation time, and the incubation location (*in situ* vs. on-deck). These challenges are further compounded in estuaries, where high temporal and spatial variability have been noted in global syntheses of photosynthesis and respiration. @cloern2014 found that annual phytoplankton primary production can vary 10-fold within an estuary and up to 5-fold from year to year. @hopkinson2005 focused on cross-system variations and found benthic and pelagic respiration rates varying by more than factors of 30 and 40, respectively, ranges that are broadly consistent with whole-system respiration rates based on the open-water method [@hoellein2013].

Advances in sensor technology for continuous collection of dissolved oxygen time series suggest that the open-water method [@Odum56] could be leveraged for applications across multiple locations. This method exploits the diel cycle of dissolved oxygen and allows the simultaneous determination of whole-system gross primary production (P $=$ net primary production $+$ autotrophic respiration), ecosystem respiration (R $=$ heterotrophic $+$ autotrophic respiration), and NEM (P $-$ R). A global synthesis of warm-season, open-water metabolism studies [@hoellein2013] was based on a total of 47 estimates, almost all of which were derived from a single study of U.S. estuaries [@caffrey2004]. A similar synthesis was done by @yvon-durocher2012. These studies evaluated the open-water method as applied to continuous monitoring data available for several decades across the National Estuarine Research Reserve System (NERRS), a network of 30 estuaries across the U.S. Thus, the approach can be broadly applied where data are available, although it is not without limitations. In particular, tidal advection can violate an assumption of the method that the measurements are from the same parcel of water with a continuous metabolic history [@kemp1980]. Striking evidence of this limitation is the large number of physically impossible negative P and R estimates in the open-water analysis of 28 3-year time series by @caffrey2003. The computed rates were negative 23% of the time, on average, and as high as 69% for individual sites.

@@

EBASE uses a mass balance equation for dissolved oxygen, assuming a well-mixed water column of depth $H$ and no lateral transport processes:

$$
H\frac{dC}{dt} = P - R - D
$$ {#eq-ebase}

where the terms on the right side of the equation are gross primary production ($P$), ecosystem respiration ($R$), and the net upward diffusive gas flux at the air--water interface ($D$), respectively, processes that result in the change in dissolved oxygen concentration ($dC$) per unit time ($dt$). The equation has dimensions of moles per unit area and time. $P$ is modeled as $aPAR$, where $PAR$ is the surface photosynthetically active radiation and $a$ is the light efficiency; $R$ is constant; and $D$ is modeled in the standard manner [e.g, @sarmiento2013] as $k_w\left(C_{sat} - C\right)$, where $k_w$ is the gas transfer velocity and $C_{sat}$ is the saturation (or equilibrium) concentration, calculated from salinity and temperature [@Garcia92]. The gas transfer velocity is modeled using the formulation in @Wanninkhof14:

@@

$$
\frac{k_w}{H} = K \cdot 1.024^{\left(T - \bar{T}\right)}
$$ {#eq-gasexbase}

@@

For ease of application, an R package was developed to implement EBASE [@Beck23; @RCT23], similar to the R package available for BASE (<https://github.com/dgiling/BASEmetab>). EBASE differs from BASE in model implementation in two ways. First, users can specify the model optimization period, where the time period can vary from a minimum of 1 day to the maximum of the entire length of the time series. This differs from the BASE approach where the optimization period is set at 1 day. Sensitivity of the EBASE method to different optimization periods is described below. Second, the EBASE software can accommodate missing observations in the input data by interpolating gaps prior to estimating the metabolic parameters. Results obtained from interpolated data are automatically removed from the output, eliminating the need to pre- and post-process the data when using EBASE. Again, this feature was not included in the original BASE method and a complete time series was required for use. A detailed web page (<https://fawda123.github.io/EBASE/>) describes how to use the package, with explanations of the functions for viewing model results, including a plot of the modeled dissolved oxygen with the observed, a time series plot of the oxygen budget terms (P, R, and D), and a time series plot of the credible intervals for the $a$, $R$, and $b$ parameters. All results provided herein were generated using the EBASE R package.

@@

The EBASE R package can also accommodate missing observations in the input data. This feature allows the estimation of metabolic rates for time series that cover relatively long periods of time (e.g., more than year), when data gaps are likely to occur for several reasons (e.g., equipment malfunction and routine maintenance). The R package accommodates data gaps using linear interpolation for any required input used by EBASE, allowing the Bayesian routine to provide continuous estimates across the time series. The results are then automatically post-processed to remove any metabolic estimates that exceed a threshold for the number of interpolated observations. The analyst can determine the appropriate length of time to use for excluding results, although the default setting for EBASE is 12 hours, where results for an entire optimization period are removed if any interpolated period exceeded 12 hours within that period. Although this does not apply to the synthetic time series, some of the estimates using the 2021 observed data for Apalachicola Bay were excluded from the analysis. This feature allows EBASE to be applied more easily to other data sets because an analyst does not need to manually screen data that include missing values prior to analysis.

>>>>

```{=tex}
\newcommand{\auni}{\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)}
\newcommand{\runi}{\text{mmol}/\text{m}^2/\text{d}}
\newcommand{\buni}{\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)}
```
*Running head*: EBASE metabolic model

@@

Estuaries play a key role in the global cycling of elements because of high rates of biogeochemical activity at the interface between land, river, ocean, and atmosphere. As materials are transported from land to rivers to ocean, estuaries profoundly transform or filter these materials through various processes, chief among which are primary production and respiration, collectively referred to as metabolism [@schubel1984]. For example, estuarine respiration of organic carbon delivered from land produces carbon dioxide, which may be lost to the atmosphere via outgassing [e.g, @cai2011]. Similarly, estuarine primary production consumes dissolved inorganic carbon delivered from land, forming particulate organic carbon that may be buried in estuarine sediments [e.g., @hu2006]. Estuarine metabolic processes are fundamentally important, as illustrated by the dissolved inorganic carbon budget of eastern North American estuaries [@najjar2018]; net primary production and heterotrophic respiration far exceed inputs from upland sources and outputs to the atmosphere and ocean. As such, small changes in either net primary production or heterotrophic respiration can have a large impact on the difference between the two, net ecosystem metabolism (NEM, sometimes also referred to as net ecosystem production), which ultimately governs the form and quantity of estuarine materials that are lost to the atmosphere (i.e., outgassing), the sediments (i.e., burial), and the ocean (i.e., advection).

Methods for estimating metabolism in aquatic environments are numerous and include open-water techniques, bottle-based incubations, ecosystem budgets, oxygen isotopes, use of inert gases, and aquatic eddy covariance [@kemp2011; @staehr2011; @berg2022]. Based on the variety of assumptions and limitations of each method, comparisons of metabolic rates within and between ecosystems have been challenging. For example, an assessment in Randers Fjord (Denmark) showed agreement in both the sign and magnitude among four different approaches [@gazeau2005]. However, gross primary production and ecosystem respiration estimates from incubations were consistently lower than those from open-water methods, consistent with the estuarine respiration synthesis of @hopkinson2005. Even within one technique, there are numerous choices to be made that alter the outcome. For example, in the commonly used ${}^{14}$C technique for measuring primary production, @cloern2014 noted that investigators have made divergent choices with regard to filtering mesozooplankton grazers, measuring ${}^{14}$C in dissolved organic carbon, changing spectral light quality with depth, the euphotic zone depth, the incubation time, and the incubation location (*in situ* vs. on-deck). These challenges are further compounded in estuaries, where high temporal and spatial variability have been noted in global syntheses of photosynthesis and respiration. @cloern2014 found that annual phytoplankton primary production can vary 10-fold within an estuary and up to 5-fold from year to year. @hopkinson2005 focused on cross-system variations and found benthic and pelagic respiration rates varying by more than factors of 30 and 40, respectively, ranges that are broadly consistent with whole-system respiration rates based on the open-water method [@hoellein2013].

Advances in sensor technology for continuous collection of dissolved oxygen time series suggest that the open-water method [@Odum56] could be leveraged for applications across multiple locations. This method exploits the diel cycle of dissolved oxygen and allows the simultaneous determination of whole-system gross primary production (P $=$ net primary production $+$ autotrophic respiration), ecosystem respiration (R $=$ heterotrophic $+$ autotrophic respiration), and NEM (P $-$ R). A global synthesis of warm-season, open-water metabolism studies [@hoellein2013] was based on a total of 47 estimates, almost all of which were derived from a single study of U.S. estuaries [@caffrey2004]. A similar synthesis was done by @yvon-durocher2012. These studies evaluated the open-water method as applied to continuous monitoring data available for several decades across the National Estuarine Research Reserve System (NERRS), a network of 30 estuaries across the U.S. Thus, the approach can be broadly applied where data are available, although it is not without limitations. In particular, tidal advection can violate an assumption of the method that the measurements are from the same parcel of water with a continuous metabolic history [@kemp1980]. Striking evidence of this limitation is the large number of physically impossible negative P and R estimates in the open-water analysis of 28 3-year time series by [@caffrey2003]. The computed rates were negative 23% of the time, on average, and as high as 69% for individual sites. Note that R is often expressed as a negative value indicating consumption of dissolved oxygen. Here and throughout, we express correct values of R as positive values given the metabolic model described below.

@@

EBASE uses a mass balance equation for dissolved oxygen, assuming a well-mixed water column of depth $Z$ and no lateral transport processes:

$$
Z\frac{dC}{dt} = P - R - D
$$ {#eq-ebase}

where the terms on the right side of the equation are gross primary production ($P$), ecosystem respiration ($R$), and the net upward diffusive gas flux at the air--water interface ($D$), respectively, processes that result in the change in dissolved oxygen concentration ($dC$) per unit time ($dt$). Both $P$ and $R$ are expressed as positive values, whereas $D$ can be positive (outgassing) or negative (ingassing). The equation has dimensions of moles per unit area and time. $P$ is modeled as $aPAR$, where $PAR$ is the surface photosynthetically active radiation and $a$ is the light efficiency; $R$ is constant; and $D$ is modeled in the standard manner [e.g, @sarmiento2013] as $k_w\left(C_{sat} - C\right)$, where $k_w$ is the gas transfer velocity and $C_{sat}$ is the saturation (or equilibrium) concentration, calculated from salinity and temperature [@Garcia92]. The gas transfer velocity is modeled using the formulation in @Wanninkhof14:

@@

$$
\frac{k_w}{Z} = K \cdot 1.024^{\left(T - \bar{T}\right)}
$$ {#eq-gasexbase}

@@

For ease of application, an R package was developed to implement EBASE [@Beck23; @RCT23], similar to the R package available for BASE (<https://github.com/dgiling/BASEmetab>). EBASE differs from BASE in model implementation in two ways. First, users can specify the model optimization period as the amount of time over which the unknown parameters are estimated from the Bayesian prior distributions (i.e., $a$, $R$, and $b$, see below). The time period can vary from a minimum of 1 day (defined as midnight to midnight the following day) to the maximum of the entire length of the time series. This differs from the BASE approach where the optimization period is set at 1 day. Sensitivity of the EBASE method to different optimization periods is described below. Note that only one estimate for each of $a$, $R$, and $b$ is returned for each optimization period as defined by the model, although $P$ and $D$ can vary continuously with each timestep given the dependence on other inputs that can vary across the time series (e.g., $PAR$ and wind). Second, the EBASE software can accommodate missing observations in the input data by interpolating gaps prior to estimating the metabolic parameters. All gaps are linearly interpolated to allow the model estimation to proceed continuously and any results for an optimization period with gaps larger than 12 hours are removed, eliminating the need to pre- and post-process the data when using EBASE. The user can vary the criteria for removing results (e.g., 6 hours as a more conservative criteria). Again, this feature was not included in the original BASE method and a complete time series was required for use. A detailed web page (<https://fawda123.github.io/EBASE/>) describes how to use the package, with explanations of the functions for viewing model results, including a plot of the modeled dissolved oxygen with the observed, a time series plot of the oxygen budget terms (P, R, and D), and a time series plot of the credible intervals for the $a$, $R$, and $b$ parameters. All results provided herein were generated using the EBASE R package.

@@

The EBASE R package can also accommodate missing observations in the input data. This feature allows the estimation of metabolic rates for time series that cover relatively long periods of time (e.g., more than year), when data gaps are likely to occur for several reasons (e.g., equipment malfunction and routine maintenance). The R package accommodates data gaps using linear interpolation for any required input used by EBASE, allowing the Bayesian routine to provide continuous estimates across the time series. The results are then automatically post-processed to remove any metabolic estimates that exceed a threshold for the number of interpolated observations. The analyst can determine the appropriate length of time to use for excluding results, although the default setting for EBASE is 12 hours, where results for an entire optimization period are removed if any interpolated period exceeded 12 hours within that period. Although this does not apply to the synthetic time series, some of the estimates using the 2021 observed data for Apalachicola Bay were excluded from the analysis. This feature allows EBASE to be applied more easily to other data sets because an analyst does not need to manually screen data that include missing values prior to analysis. More conservative gaps can be chosen if, for example, semidiurnal tidal signals operating at time scales less than 12 hours are expected to influence the results. However, detiding is strongly encouraged in these cases and the default gap size of 12 hours that describes the diurnal oxygen signal is likely appropriate after detiding.
